Did WW2 Produce an Equivalent to a Modern Computer?

### 2017-06-18 Draft 2/Appendices.tex
Hmm looks like we're all done here
WIth 17 hours 42 minutes to go as well, I'm impressed

\begin{document}
\appendix
\appendixpage
\section{Further Arithmetic in $\lambda$-Calculus}
Multiplication can also be calculated by defining the function \textbf{M}:
\cite{rojas2015lambdatutorial}
\begin{equation*}
\bm{\mathrm{M}} = \lambda wyx.w(yx)
\end{equation*}  
\begin{gather*}
\begin{aligned}
\bm{\mathrm{M22}} &= (\lambda wyx.w(yx))(\lambda ab.a(a(b)))(\lambda sz.s(s(z)))\\
&= \lambda x.(\lambda ab.a(a(b)))[(\lambda sz.s(s(z)))x]\\
&= \lambda x.(\lambda ab.a(a(b)))[\lambda z.x(x(z))]\\
&= \lambda x.(\lambda b.[\lambda z.x(x(z))]([\lambda z.x(x(z))]b))\\
&= \lambda x.(\lambda b.[\lambda z.x(x(z))][x(x(b))])\\
&= \lambda x.(\lambda b.x(x(x(x(b)))))\\
&= \lambda xb.x(x(x(x(b))))\\
&= \bm{\mathrm{4}}
\end{aligned}
\end{gather*}

An `exponent' function to find $a^b$can also be defined, of astonishing simplicity:
\cite{barendregt1984introduction}
\begin{equation*}
\bm{\mathrm{E}} = \lambda ba.ab
\end{equation*}
\begin{gather*}
\begin{aligned}
\bm{\mathrm{E23}} &= (\lambda ab.ab)(\lambda sz.s(s(z)))(\lambda ty.t(t(t(y))))\\
&= (\lambda sz.s(s(z)))(\lambda ty.t(t(t(y))))\\
&= (\lambda z.[\lambda ty.t(t(t(y)))]([\lambda ux.u(u(u(x)))]z))\\
&= (\lambda z.[\lambda ty.t(t(t(y))][\lambda x.z(z(z(x)))])\\
&= (\lambda z.[\lambda y.[\lambda x.z(z(z(x)))]([\lambda x.z(z(z(x)))]([\lambda x.z(z(z(x)))]y))])\\
&= (\lambda z.[\lambda y.[\lambda x.z(z(z(x))]([\lambda x.z(z(z(x)))][z(z(z(y)))])])\\
&= (\lambda z.[\lambda y.[\lambda x.z(z(z(x))]([z(z(z(z(z(z(y))))))])])\\
&= (\lambda z.[\lambda y.[z(z(z(z(z(z(z(z(z(y)))))))))]])\\
&= \lambda zy.z(z(z(z(z(z(z(z(z(y)))))))))\\
&= \bm{\mathrm{9}} = \bm{\mathrm{3^2}}
\end{aligned}
\end{gather*}

@@

Producing the sequence \texttt{010110111011110111110...} on primary squares, this machine operates by first printing the symbols \texttt{e e 0} at the start of the tape, before printing the first \texttt{0} of the sequence on the fifth cell, finishing on the third cell and moving to state $\mathbb{A}$.

@@

\section{Detailed Demonstration of the Universality of $\lambda$-Calculus}\label{appendix:lambdauniversaldemo}

@@

\medskip\noindent\begin{tabu} to \textwidth{XXXX}
    \multicolumn{2}{c}{\textit{Configuration}} & \multicolumn{2}{c}{\textit{Behavior}} \\
    \textit{state} & \textit{symbol} & \textit{transition function} & \textit{final state} \\
    \hhline{====}
    $q_i$ & $S_i$ & P$S_k$, $M_{L/R/S}$ & $q_m$ \\
    \hhline{----}
\end{tabu}

@@

\noindent\begin{tabu} to \textwidth{XXXX}
    $q_i$ & $S_i$ & E, $M_{L/R/S}$ & $q_m$ \\
\end{tabu}
\medskip

@@

\noindent\begin{tabu} to \textwidth{XXXX}
    $q_i$ & $S_i$ & P$S_0$, $M_{L/R/S}$ & $q_m$ \\
\end{tabu}
\medskip

@@

\noindent\begin{tabu} to \textwidth{XXXX}
    $q_i$ & $S_i$ & $M_{L/R/S}$ & $q_m$ \\
\end{tabu}
\medskip

@@

\noindent\begin{tabu} to \textwidth{XXXX}
    $q_i$ & $S_i$ & P$S_i$, $M_{L/R/S}$ & $q_m$ \\
\end{tabu}
\medskip

\noindent in order for all lines to be written in the first format. As the number of symbols used for a machine is always finite, it is possible to enumerate under every single state an action for every single symbol $S_i$ that the machine could be reading. As such, `All' and `Not \textit{a}' may be portrayed in this format. This idea also allows the last simplification shown in the table above to work.

@@

\subsection{Emulating the Tape}\label{sec:LambdaUnivTape}
To allow for storage of large sequences of data in $\lambda$-Calculus, ordered lists must be defined. These will be used to store the content of the tape. These are defined as a sequence of nodes, where each node is a pair with the first element as the content of the node and the second element is the next node in the sequence (as shown in \cite{tromp2007binary}). The final node will have \textbf{0} as its first element to signify the end of the list - the second element will never be referred to; as such \textbf{0} will be used as a placeholder (of course this only works if the list is not designed to contain any values of zero or true, as they will be indistinguishable from the end marker - I shall use \textbf{1} to represent $S_0$, \textbf{2} to represent $S_1$ and so on). For example, the list [a,b,c,d,e] would be represented as:
\begin{equation*}
\lambda l.la(\lambda m.mb(\lambda n.nc(\lambda o.od(\lambda p.pe(\lambda z.z\bm{\mathrm{00}})))))
\end{equation*}
In order to obtain the first element (head), the argument \textbf{T} would be applied to the list; to get the remaining elements (tail) the \textbf{F} argument would be applied. 
Furthermore, to prepend an element $x$ to the start of a list $l$:
\begin{equation*}
\bm{\mathrm{:}} = \lambda xlf.f[x][l]
\end{equation*}

The tape of the machine extends infinitely in both directions either side of the cursor. This shal be represented as a pair of lists: $T_p$ as the list of all the predecessors to the cursor and $T_s$ as the list of all the successors to the cursor. The predecessors shall be listed in reverse order, such that the head of $T_p$ is the symbol directly to the left of the cursor. The head of $T_s$ is the symbol considered to be `under' the cursor, the next element is the symbol to the right of that, and so on.
\begin{equation*}
T_a = \lambda f.f T_p T_s
\end{equation*}

@@

Unfortunately, the tape which that is currently being dealt with is unlike that of a Turing machine, due to the fact that it is of finite length - one cannot write a single list in $\lambda$-Calculus which expresses the entire tape. As such, when the machine moves left or right onto the node with \textbf{0} as the head signifying the end of the list, it shall be replaced with a node containing the blank square symbol $S_0$, represented by a \textbf{1}, as the head, and another end marker node as the tail. To do this, another transformation function shall be used which will check if either head element of the tape pair is \textbf{0}, and if so, will replace it with an empty square node $\bm{\epsilon}$ - otherwise it will simply be replaced with the original head:

@@

In order to find which action should be performed given a certain configuration, one can use the following function which produces a pair of a list transition function and a state reference from a current state, the current symbol, and the description function. As part of this function, the recursive function $\bm{\rho}$ must be used in conjunction with the \textbf{Y} combinator, which returns the action aspect of the first pair under symbol $s$ in list $l$ (the $r$ argument is the argument which would represent the function as a whole, and is recursively called by the \textbf{Y} combinator)

>>>>

\begin{document}
\appendix
\appendixpage
\section{Further Arithmetic in $\lambda$-Calculus}\label{appendix:lambdaarithmetic}

\subsection{Exponents}

An `exponent' function to find $a^b$ of astonishing simplicity can be defined:
\cite{barendregt1984introduction}
\begin{equation*}
\bm{\mathrm{E}} = \lambda ab.ba
\end{equation*}
\begin{gather*}
\begin{aligned}
\bm{\mathrm{E32}} &= (\lambda ab.ba)\ (\lambda sz.s(s(s(z))))\ (\lambda ty.t(t(y)))\\
&= \lambda ty.t(t(y)))\ (\lambda sz.s(s(s(z)))\\
&= \lambda y.[\lambda sz.s(s(s(z)))]\ ([\lambda ux.u(u(u(x)))]\ y)\\
&= \lambda y.[\lambda sz.s(s(s(z)))]\ [\lambda x.y(y(y(x)))]\\
&= \lambda y.[\lambda z.[\lambda x.y(y(y(x)))]\ ([\lambda x.y(y(y(x)))]\ ([\lambda x.y(y(y(x)))]\ z))]\\
&= \lambda y.[\lambda z.[\lambda x.y(y(y(x)))]\ ([\lambda x.y(y(y(x)))]\ [y(y(y(z)))])]\\
&= \lambda y.[\lambda z.[\lambda x.y(y(y(x))]\ ([y(y(y(y(y(y(z))))))])]\\
&= \lambda y.[\lambda z.[y(y(y(y(y(y(y(y(y(z)))))))))]]\\
&= \lambda yz.y(y(y(y(y(y(y(y(y(z)))))))))\\
&= \bm{\mathrm{9}} = \bm{\mathrm{3^2}}
\end{aligned}
\end{gather*}

\subsection{Integers}

Integers are defined as a pair of natural numbers, whose difference is the value of the integer (any function that outputs an integer shall be written with a bar):
\begin{gather*}
\begin{aligned}
\lvari{-1} &= \lambda x.x\ \lvar{01} &= \lambda x.x\ \lvar{12}\\
\lvari{-2} &= \lambda x.x\ \lvar{02} &= \lambda x.x\ \lvar{13}
\end{aligned}    
\end{gather*}

\begin{gather*}
\begin{aligned}
\lvari{S} &= \lambda nx.x\ [\lvar{S}(n\lvar{T})][n\lvar{F}] \\
\lvari{P} &= \lambda nx.x\ [n\lvar{T}][\lvar{S}(n\lvar{F})]
\end{aligned}    
\end{gather*}

\begin{gather*}
\begin{aligned}
\lvari{A} &= \lambda abx.x\ [(a\lvar{T})\lvar{S}(b\lvar{T})]\ [(a\lvar{F})\lvar{S}(b\lvar{F})] \\
\lvari{D} &= \lambda abx.x\ [(a\lvar{T})\lvar{S}(b\lvar{F})]\ [(a\lvar{F})\lvar{S}(b\lvar{T})] \\
\lvari{M} &= \lambda abx.x\ [(\lvar{M}\ [a\lvar{T}][b\lvar{T}])\ \lvar{S}\ (\lvar{M}\ [a\lvar{F}][b\lvar{F}])]\ [(\lvar{M}\ [a\lvar{T}][b\lvar{F}])\ \lvar{S}\ (\lvar{M}\ [a\lvar{F}][b\lvar{T}])] \\
\end{aligned}    
\end{gather*}

The \textbf{$\pm$} function outputs \textbf{T} for positive integers and \textbf{F} for negative integers ($\lvari{0}$ is considered to be positive), and the \textbf{$|$} function calculates the absolute value of the given integer.

\begin{gather*}
\begin{aligned}
\lvar{\pm} &= \lambda i.(\lvar{\geq}\ [i\lvar{T}]\ [i\lvar{F}]) \\
\lvar{|} &= \lambda i. (\lvar{\pm}i)\ [D\ (i\lvar{T}) (i\lvar{F})]\ [D\ (i\lvar{F}) (i\lvar{T})]
\end{aligned}    
\end{gather*}

Rational numbers can be represented in a similar way as a fraction of two integers.

\subsection{Division of Natural Numbers}

The \textbf{Q} function (which of course stands for quotient) calculates the number of times $a$ is contained within $b$ recursively: \cite{allisonWEB17integers}

\begin{gather*}
\begin{aligned}
\lvar{\delta} &= \lambda dpq.(\lvar{\geq}pq)\ [\lvar{S}\ (d\ [\lvar{D}pq]\ q)]\ [\lvar{0}] \\
\lvar{Q} &= \lambda ab.\lvar{Y\delta}ab
\end{aligned}    
\end{gather*}

@@

Producing the sequence \texttt{010110111011110111110...} on primary squares \cite{turing1936computablenumbers}, this machine operates by first printing the symbols \texttt{e e 0} at the start of the tape, before printing the first \texttt{0} of the sequence on the fifth cell, finishing on the third cell and moving to state $\mathbb{A}$.

@@

\section{Demonstration of the Universality of $\lambda$-Calculus}\label{appendix:lambdauniversaldemo}

@@

\medskip\noindent\begin{tabu} to \textwidth{XXXX}
    \multicolumn{2}{c}{\textit{Configuration}} & \multicolumn{2}{c}{\textit{Behavior}} \\
    \textit{state} & \textit{symbol} & \textit{transition function} & \textit{final state} \\
    \hhline{====}
    $q_i$ & $S_j$ & P$S_k$, $M_{L/R/S}$ & $q_m$ \\
    \hhline{----}
\end{tabu}

@@

\noindent\begin{tabu} to \textwidth{XXXX}
    $q_i$ & $S_j$ & E, $M_{L/R/S}$ & $q_m$ \\
\end{tabu}
\medskip

@@

\noindent\begin{tabu} to \textwidth{XXXX}
    $q_i$ & $S_j$ & P$S_0$, $M_{L/R/S}$ & $q_m$ \\
\end{tabu}
\medskip

@@

\noindent\begin{tabu} to \textwidth{XXXX}
    $q_i$ & $S_j$ & $M_{L/R/S}$ & $q_m$ \\
\end{tabu}
\medskip

@@

\noindent\begin{tabu} to \textwidth{XXXX}
    $q_i$ & $S_j$ & P$S_j$, $M_{L/R/S}$ & $q_m$ \\
\end{tabu}
\medskip

\noindent in order for all lines to be written in the first format. As the number of symbols used for a machine is always finite, it is possible to enumerate under every single state an action for every single symbol $S_j$ that the machine could be reading. As such, `All' and `Not \textit{a}' may be portrayed in this format. This idea also allows the last simplification shown in the table above to work.

@@

\subsection{Emulating the Tape}\label{sec:LambdaUnivTape}
To allow for storage of large sequences of data in $\lambda$-calculus, ordered lists must be defined. These will be used to store the content of the tape. These are defined as a sequence of nodes, where each node is a pair with the first element as the content of the node and the second element is the next node in the sequence (as shown in \cite{tromp2007binary}). The final node will have \textbf{0} as its first element to signify the end of the list - the second element will never be referred to; as such \textbf{0} will be used as a placeholder (of course this only works if the list is not designed to contain any values of zero or true, as they will be indistinguishable from the end marker - I shall use \textbf{1} to represent $S_0$, \textbf{2} to represent $S_1$ and so on). For example, the list [a,b,c,d,e] would be represented as:
\begin{equation*}
\lambda l.la(\lambda m.mb(\lambda n.nc(\lambda o.od(\lambda p.pe(\lambda z.z\bm{\mathrm{00}})))))
\end{equation*}   
In order to obtain the first element (head), the argument \textbf{T} would be applied to the list; to get the remaining elements (tail) the \textbf{F} argument would be applied. 
Furthermore, to prepend an element $x$ to the start of a list $l$:
\begin{equation*}
\bm{\mathrm{:}} = \lambda xlf.f[x][l]
\end{equation*}

The tape of the machine extends infinitely in both directions either side of the cursor. This shall be represented as a pair of lists: $T_p$ as the list of all the predecessors to the cursor and $T_s$ as the list of all the successors to the cursor. The predecessors shall be listed in reverse order, such that the head of $T_p$ is the symbol directly to the left of the cursor. The head of $T_s$ is the symbol considered to be `under' the cursor, the next element is the symbol to the right of that, and so on.
\begin{equation*}
T_a = \lambda f.f T_p T_s
\end{equation*}

@@

Unfortunately, the tape which that is currently being dealt with is unlike that of a Turing machine, due to the fact that it is of finite length - one cannot write a single list in $\lambda$-calculus which expresses the entire tape. As such, when the machine moves left or right onto the node with \textbf{0} as the head signifying the end of the list, it shall be replaced with a node containing the blank square symbol $S_0$, represented by a \textbf{1}, as the head, and another end marker node as the tail. To do this, another transformation function shall be used which will check if either head element of the tape pair is \textbf{0}, and if so, will replace it with an empty square node $\bm{\epsilon}$ - otherwise it will simply be replaced with the original head:

@@

In order to find which action should be performed given a certain configuration, one can use the following function which produces a pair of a list transition function and a state reference from a current state $q$, the current symbol $s$, and the description function $d$
. As part of this function, the recursive function $\bm{\rho}$ must be used in conjunction with the \textbf{Y} combinator, which returns the action aspect of the first pair under symbol $s$ in list $l$ (the $r$ argument is the argument which would represent the function as a whole, and is recursively called by the \textbf{Y} combinator).

### 2017-06-18 Draft 2/Master.tex
Hmm looks like we're all done here
WIth 17 hours 42 minutes to go as well, I'm impressed

   \documentclass {article}
\usepackage[a4paper, total={6in, 8in}]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{subfiles}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{appendix}
\usepackage{blindtext}
\usepackage{bm}
\usepackage{cleveref}
\usepackage{hhline}
\usepackage{listings}
\usepackage[bb=boondox]{mathalfa}
\usepackage{multirow, bigdelim}
\usepackage{tabu}
\usepackage{url}
\usepackage{yfonts}

@@

\renewcommand{\arraystretch}{1.5}

\begin{document}
\pagenumbering{roman}
\maketitle
%\tableofcontents
\medskip

\begin{abstract}
Computability, or the potential for a device to solve any solvable problem given to it, is an informal notion. Its definition can be refined with use of the $\lambda$-Calculus or Turing machines, theoretical computational devices which while having simple definitions produce functionality similar to that of a computer. Hence, a computer can be defined as a machine which emulates either of these devices fully. It will then be shown that the Z3 computer built by Konrad Zuse implements computability with some success.
\end{abstract}

@@

The most significant problem in answering this question is the actual definition of `computer'. In the early 20th Century, prior to developments in this field, any human which evaluated statements through computation was known, as a `computer'. We will therefore explore the qualities that an entity - human, machine or other - must have for it to be able to perform computation.

The 18th Century German polymanth Gottfreid Leibniz proposed the ideal of a `Characteristica Universalis', a universal language in which any possible problem can be expressed, and a decision method which could evaluate any problem in this language \cite{barendregt1984introduction}. This idea is a final aim for computation - a system which when given a statement in this universal language is able to evaluate it. The ideal of producing a simple system which can evaluate any `effectively computable' statment is the main challenge of computation.

Therefore, in order for a machine to be called a `computer', it must be able to evaluate any member of a set of `computable functions' (problems expressed in the universal tongue). However, this is an abstract notion - to clarify it, a simple language in which we can express computable functions must be defined, and a simple method to evaluate them must be found. The two most common methods, $\lambda$-Calculus and the Turing Machine, shall be explored and evaluated in terms of their similarities and differences, before they are used to evaluate early computers.

@@

In our journey, we began with the abstract philosophical concept of computability, knowing nothing beyond roughly what we aim to achieve with it. Defining $\lambda$-Calculus, we were able to see that computation in all its power follows simply from the very natural concept of functions. Defining Turing machines, we saw that this abstract concept is not distant from a physical realisation; and exploring the functionality of Z3, that even relatively simplistic machines can demonstrate computation. It has become clear that the target of computability is not as difficicult to achive as it could appear, and that it was achived before the end of World War 2.

The fact that we have been able to provide a unified definition of logical decision-making along the way is, in my opinion, the most remarkable result of this study. We must consider, however, that there are limits to computation - firstly, there is a set of so called 'undecideable' problems with solutions which cannot be solved algorithmically, including finding integer solutions to Diophantine equations \cite{matiyasevich1993hilbert} and the Halting problem, which would determine whether a given computer program runs indefinitely \cite{turing1936computablenumbers}. Furthermore, when the solution to a problem is probabalistic (e.g. in simulation of quantum processes - see \cite{feynman1982simulating}) as all computation produces the same solution given a problem. As such, despite effective computability being a simple, beautiful and practical theory, it does not achieve Leibniz's aim of a Universal Language; in truth, I believe that the existance of an algorithmic description of everything would be a surprising result. Of course, Mathematics is filled with surprises.

\subfile{"Appendices"}

\bibliography{bibliography}
\bibliographystyle{unsrt}
\end{document}

>>>>

\documentclass {article}
\usepackage[a4paper, total={6in, 8in}, top=1.3in, bottom=1.5in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{subfiles}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{appendix}
\usepackage{blindtext}
\usepackage{bm}
\usepackage{cleveref}
\usepackage{hhline}
\usepackage{listings}
\usepackage[bb=boondox]{mathalfa}
\usepackage{multirow, bigdelim}
\usepackage{tabu}
\usepackage{url}
\usepackage{yfonts}

@@

\renewcommand{\arraystretch}{1.5}
\newcommand*\lvar[1]{\bm{\mathrm{#1}}}
\newcommand*\lvari[1]{\overline{\bm{\mathrm{#1}}}}

\begin{document}
\pagenumbering{roman}
\maketitle
%\tableofcontents
\medskip
\medskip

\begin{abstract}
Computability, or the potential for a device to solve any solvable problem given to it, is an informal notion. Its definition can be refined with use of the $\lambda$-calculus or Turing machines, theoretical computational devices which while having simple definitions produce functionality similar to that of a computer. Hence, a computer can be defined as a machine which emulates either of these devices fully. It will then be shown that the Z3 computer built by Konrad Zuse during World War 2 implements computability with some success.
\end{abstract}

@@

The most significant problem in answering this question is the actual definition of `computer'. In the early 20th Century, prior to developments in this field, any human who evaluated statements through computation was known as a `computer'. We will therefore explore the qualities that an entity - human, machine or other - must have for it to be able to perform computation.

The 18th Century German polymath Gottfreid Leibniz proposed the ideal of a `Characteristica Universalis', a universal language in which any possible problem can be expressed, and a decision method which could evaluate any problem in this language \cite{barendregt1984introduction}. This idea is a final aim for computation - a system which when given a statement in this universal language is able to evaluate it. The ideal of producing a simple system which can evaluate any `effectively computable' statement is the main challenge of computation.

Therefore in order for a machine to be called a `computer' it must be able to evaluate any member of a set of `computable functions' (problems expressed in the universal tongue). However, this is an abstract notion - to clarify it, a simple language in which we can express computable functions must be defined, and a simple method to evaluate them must be found. The two most common methods, $\lambda$-
and the Turing machine, shall be explored, before they are used to evaluate early computers.

Throughout this paper any equation, table or code listing which is not clearly cited can be considered my own work.

\clearpage

@@

In our journey, we began with the abstract philosophical concept of computability, knowing nothing beyond roughly what we aim to achieve with it. Defining $\lambda$-calculus, we were able to see that computation in all its power follows simply from the very natural concept of functions. Defining Turing machines, we saw that this abstract concept is not distant from a physical realisation; and exploring the functionality of Z3 we found that even relatively simplistic machines can demonstrate computation. It has become clear that the target of computability is not as difficult to achieve as it could appear, and that it was achieved before the end of World War 2.

The fact that we have been able to provide a unified definition of logical decision-making along the way is, in my opinion, the most remarkable result of this study. We must consider, however, that there are limits to computation - firstly, there is a set of so called `undecideable' problems with solutions which cannot be solved algorithmically, including finding integer solutions to Diophantine equations \cite{matiyasevich1993hilbert} and the Halting problem, which would determine whether a given computer program runs indefinitely \cite{turing1936computablenumbers}. Furthermore, computation fails when the solution to a problem is probabilistic (e.g. in simulation of quantum processes - see \cite{feynman1982simulating}) as all algorithms produce the same solution given a problem. As such, despite effective computability being a simple, beautiful and practical theory, it does not achieve Leibniz's aim of a Universal Language; in truth, I believe that the existence of an algorithmic description of everything would be a surprising result. Of course, Mathematics is filled with surprises.

\subfile{"Appendices"}
\clearpage
\bibliography{bibliography}
\bibliographystyle{unsrt}
\end{document}
